**The status of the microbial census: an update**

### 2015-11-25 Schloss_Census2_mBio_2015.Rmd
Fix r code in M&M section

***Sequence data curation.***
The July 19, 2015 release of the ARB-formatted SILVA small subunit (SSU) reference database (SSU Ref v.123) was downloaded from http://www.arb-silva.de/fileadmin/silva_databases/release_123/ARB_files/SSURef_123_SILVA_19_07_15_opt.arb.tgz [@REF]. This release is based on the EMBL-EBI/ENA Release 123, which was released in March 2015. The SILVA curators identify potential SSU sequences using keyword searches and sequence-based search using RNAmmer (http://www.arb-silva.de/documentation/release-123/). The SILVA curators then screened the 7,168,241 resulting sequences based on a minimum length criteria (<300 nt), number of ambiguous base calls (>2%), length of sequence homopolymers (>2%), presence of vector contamination (>2%), low alignment quality value (<75), and likelihood of being chimeric (Pintail value < 50). Of the remaining sequences, the bacterial reference set retained those bacterial sequences longer than 1,200 nt and the archaeal reference set retained those archaeal sequences longer than 900 nt. The aligned 1,515,024 bacterial and 59,240 archaeal sequences were exported from the database using ARB along with the complete set of metadata. Additional sequence data was included from single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], many of which were recently obtained via the GEBA-MDM effort in Rinke et al. [@REF]. "SCGC” was searched on the IMG database March 12, 2015 to download the bacterial (N=249) and archaeal (N=46) 16S rRNA gene sequences and their associated metadata. The IMG sequences were aligned against the respective SILVA-based reference using mothur [@REF]. The aligned bacterial and archaeal sequence sets from SILVA and IMG were pooled and processed in parallel in mothur. Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates. The sequences in the resulting bacterial dataset overlapped bases 113 through 1344 of an *E. coli* reference sequence (V00348) and had a median length of 1,227 nt. The sequences in the resulting archaeal dataset overlapped positions 362 to 937 of a *Sulfolobus solfataricus* reference sequence (X03235) and had a median length of 580 nt. The archaeal sequences were considerably shorter than their initial length because it was necessary to find a common overlapping region across the sequences. The final datasets contained `r nrow(bact)` bacterial and `r nrow(arch)` archaeal 16S rRNA gene sequences. Sequences were assigned to OTUs using the average neighbor clustering algorithm [@REF].

***Metadata curation.***
The metadata that was contained within the SSU Ref database was used to expand our analysis beyond a basic count of sequences and the number of OTUs in each domain. The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse" categories, and further refined to facilitate additional analyses with twenty-six more specific "fine" categories (Table 1). These were assigned based on manual curation of the "isolation_source" category within the ARB database associated with each of the sequences. There were `r length(unique(bact$isolation_source))-1` bacterial and `r length(unique(arch$isolation_source))-1` archaeal isolation_source terms that were manually assigned to the coarse and fine categories in Table 1. For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category. There were `r sum(is.na(bact$isolation_source))` bacterial and `r sum(is.na(arch$isolation_source))` archaeal sequences where an "isolation_source" term was not collected. We ascertained whether a sequence came from a cultured organism by including those sequences that had data in their "strain" or "isolate" fields within the database and excluded any sequences that had "Unc" as part of their database name as this is a convention in the database that represents sequences from uncultured organisms. Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our categorization are available at [@figshare].

@@

b_sobs <- get_sobs(read.table(file='data/mothur/all_bacteria.filter.unique.precluster.an.rarefaction', header=T))
a_sobs <- get_sobs(read.table(file='data/mothur/all_archaea.filter.unique.precluster.an.rarefaction', header=T))

category_summary <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T)
b_goods <- category_summary["total", "bact_good_coverage"]
b_schloss <- category_summary["total", "bact_schloss_coverage"]

a_goods <- category_summary["total", "arch_good_coverage"]
a_schloss <- category_summary["total", "arch_schloss_coverage"]
```

***The status of the bacterial and archaeal census.***
To assess the field's progress in characterizing the biodiversity of bacteria and archaea we assigned each 16S rRNA gene sequence to OTUs using distance threshold varying between 0 and 20%. Although it is not possible to link a specific taxonomic level (e.g. species, genus, family, etc.) to a specific distance threshold, we selected distances of 0, 3, 5, 10, and 20% because they are widely regarded as representing the range of genetic diversity of the 16S rRNA gene within each domain. By rarefaction, it was clear that the ongoing sampling efforts have started to saturate the number of current OTUs. After sampling `r nrow(bact)` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the selected thresholds (Figure 1A, Table 1). Using on the OTUs generated using a 3% threshold, we found that `r b_goods`% of the sequences belonged to OTUs that had been observed more than once. In contrast, only `r b_schloss`% of the OTUs that were observed had been seen more than once. Paralleling the bacterial results, after sampling `r nrow(arch)` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure 1B, Table 1). Using on the OTUs generated using a 3% threshold, we found that `r a_goods`% of the sequences belonged to OTUs that had been observed more than once. In contrast, only `r a_schloss`% of the OTUs that were observed had been seen more than once. The results for the bacterial and archaeal censuses indicate that regardless of the domain, if we continue sampling with the current strategies we will continue to sample OTUs that have already been observed even though a large fraction of OTUs have only been sampled once. Furthermore, considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once it is likely that an even larger number of OTUs have yet to be sampled for both domains.

@@

```{r taxon_census}
	b_phylum_year <- table(bact_year_deposited, bact_phylum)
	b_phylum_n <- apply(b_phylum_year, 2, sum)
	b_n_phylum_2015 <- ncol(b_phylum_year)
	b_phylum_2006 <- b_phylum_year[rownames(b_phylum_year) %in% as.character(1983:2006),]
	b_n_phylum_2006 <- sum(apply(b_phylum_2006, 2, sum) > 0) #how many phyla were there?
	b_n_old_seqs <- sum(b_phylum_2006)						#how many seqs in 2006?
	b_n_old_seqs_phyla <- apply(b_phylum_2006, 2, sum)		#how many sequences in each phyla?

>>>>

```{r overview}
category_summary <- read.table(file="data/process/coverage_by_category_and_time.tsv", header=T)

b_nseqs <- category_summary["total", "bact_count"]
b_goods <- category_summary["total", "bact_good_coverage"]
b_schloss <- category_summary["total", "bact_schloss_coverage"]

a_nseqs <- category_summary["total", "arch_count"]
a_goods <- category_summary["total", "arch_good_coverage"]
a_schloss <- category_summary["total", "arch_schloss_coverage"]
```

***Sequence data curation.***
The July 19, 2015 release of the ARB-formatted SILVA small subunit (SSU) reference database (SSU Ref v.123) was downloaded from http://www.arb-silva.de/fileadmin/silva_databases/release_123/ARB_files/SSURef_123_SILVA_19_07_15_opt.arb.tgz [@REF]. This release is based on the EMBL-EBI/ENA Release 123, which was released in March 2015. The SILVA curators identify potential SSU sequences using keyword searches and sequence-based search using RNAmmer (http://www.arb-silva.de/documentation/release-123/). The SILVA curators then screened the 7,168,241 resulting sequences based on a minimum length criteria (<300 nt), number of ambiguous base calls (>2%), length of sequence homopolymers (>2%), presence of vector contamination (>2%), low alignment quality value (<75), and likelihood of being chimeric (Pintail value < 50). Of the remaining sequences, the bacterial reference set retained those bacterial sequences longer than 1,200 nt and the archaeal reference set retained those archaeal sequences longer than 900 nt. The aligned 1,515,024 bacterial and 59,240 archaeal sequences were exported from the database using ARB along with the complete set of metadata. Additional sequence data was included from single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], many of which were recently obtained via the GEBA-MDM effort in Rinke et al. [@REF]. "SCGC” was searched on the IMG database March 12, 2015 to download the bacterial (N=249) and archaeal (N=46) 16S rRNA gene sequences and their associated metadata. The IMG sequences were aligned against the respective SILVA-based reference using mothur [@REF]. The aligned bacterial and archaeal sequence sets from SILVA and IMG were pooled and processed in parallel in mothur. Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates. The sequences in the resulting bacterial dataset overlapped bases 113 through 1344 of an *E. coli* reference sequence (V00348) and had a median length of 1,227 nt. The sequences in the resulting archaeal dataset overlapped positions 362 to 937 of a *Sulfolobus solfataricus* reference sequence (X03235) and had a median length of 580 nt. The archaeal sequences were considerably shorter than their initial length because it was necessary to find a common overlapping region across the sequences. The final datasets contained `r b_nseqs` bacterial and `r a_nseqs` archaeal 16S rRNA gene sequences. Sequences were assigned to OTUs using the average neighbor clustering algorithm [@REF].

***Metadata curation.***
The metadata that was contained within the SSU Ref database was used to expand our analysis beyond a basic count of sequences and the number of OTUs in each domain. The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse" categories, and further refined to facilitate additional analyses with twenty-six more specific "fine" categories (Table 1). These were assigned based on manual curation of the "isolation_source" category within the ARB database associated with each of the sequences (Table 1). For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category. There were `r category_summary["no_source", "bact_count"]` bacterial and `r category_summary["no_source", "arch_count"]` archaeal sequences where an "isolation_source" term was not collected. We ascertained whether a sequence came from a cultured organism by including those sequences that had data in their "strain" or "isolate" fields within the database and excluded any sequences that had "Unc" as part of their database name as this is a convention in the database that represents sequences from uncultured organisms. Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our environemtnal categorizations are available at [@figshare].

@@

b_sobs <- get_sobs(read.table(file='data/mothur/all_bacteria.filter.unique.precluster.an.rarefaction', header=T))
a_sobs <- get_sobs(read.table(file='data/mothur/all_archaea.filter.unique.precluster.an.rarefaction', header=T))
```

***The status of the bacterial and archaeal census.***
To assess the field's progress in characterizing the biodiversity of bacteria and archaea we assigned each 16S rRNA gene sequence to OTUs using distance threshold varying between 0 and 20%. Although it is not possible to link a specific taxonomic level (e.g. species, genus, family, etc.) to a specific distance threshold, we selected distances of 0, 3, 5, 10, and 20% because they are widely regarded as representing the range of genetic diversity of the 16S rRNA gene within each domain. By rarefaction, it was clear that the ongoing sampling efforts have started to saturate the number of current OTUs. After sampling `r b_nseqs` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the selected thresholds (Figure 1A, Table 1). Using on the OTUs generated using a 3% threshold, we found that `r b_goods`% of the sequences belonged to OTUs that had been observed more than once. In contrast, only `r b_schloss`% of the OTUs that were observed had been seen more than once. Paralleling the bacterial results, after sampling `r nrow(arch)` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure 1B, Table 1). Using on the OTUs generated using a 3% threshold, we found that `r a_goods`% of the sequences belonged to OTUs that had been observed more than once. In contrast, only `r a_schloss`% of the OTUs that were observed had been seen more than once. The results for the bacterial and archaeal censuses indicate that regardless of the domain, if we continue sampling with the current strategies we will continue to sample OTUs that have already been observed even though a large fraction of OTUs have only been sampled once. Furthermore, considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once it is likely that an even larger number of OTUs have yet to be sampled for both domains.

@@

##### here #######
```{r taxon_census}
	b_phylum_year <- table(bact_year_deposited, bact_phylum)
	b_phylum_n <- apply(b_phylum_year, 2, sum)
	b_n_phylum_2015 <- ncol(b_phylum_year)
	b_phylum_2006 <- b_phylum_year[rownames(b_phylum_year) %in% as.character(1983:2006),]
	b_n_phylum_2006 <- sum(apply(b_phylum_2006, 2, sum) > 0) #how many phyla were there?
	b_n_old_seqs <- sum(b_phylum_2006)						#how many seqs in 2006?
	b_n_old_seqs_phyla <- apply(b_phylum_2006, 2, sum)		#how many sequences in each phyla?
